The tree line ecotone punctuates the elevational turnover of plants, pollinators, and their interactions

### 2021-04-06 ms_1/ms_1.Rmd
added management covariate to models

---
title: "The tree line ecotone punctuates the elevational turnover of plants, pollinators, and their interactions "

author: Douglas B. Sponsler^1^*, Fabrice Requier^2^, Katharina Kallnik^1^, Alice Claßen^1^, A. Fabienne Maihoff^1^, Ingolf Steffan-Dewenter^1^

@@

@@

@@

The structuring of biological communities along mountain slopes is complex, and the elevational range shifts that are now occurring in response to climate change involve more than merely tracking suitable temperature envelopes. When species move, they do so in the context of biological communities, and the outcomes of these movements depend on how and to what extent biotic interactions are reordered. Bumble bees (Hymentopera: *Bombus* spp.) are cold-adapted species associated with mountain habitats, and they have already exhibited measurable upslope range shifts that are expected to result in habitat loss, novel competitive interactions, and the rewiring of pollination networks. Predicting and interpreting these shifts, however, requires an understanding of the current elevational patterns of bumble bees and the floral mutualists that are being acted upon by climate change. In this study, we sampled bumble bees, flowering plants, and their interactions over three consecutive years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling techniques, we analyze the elevational patterns of this community at three levels of biological organization: species abundance, species β-diversity, and interaction β-diversity. We demonstrate complex, nonlinear responses to elevation at all three levels of organization. In particular, we identify the tree line ecotone as (1) a distributional interface between low/mid- and high-elevation bumble bee species, (2) a threshold above which floral resource availability sharply decreases, and (3) a zone of accelerated turnover of flowering plants and bumble- bee-flower interactions. The implications of these findings extend beyond the particular case of bumble bees to demonstrate that linear elevational temperature gradients are ecologically punctuated, and the outcomes of climate-induced range shifts with depend on how species and species-interactions change as they cross --- or fail to cross --- thresholds like the tree line ecotone.

*Keywords*: climate, elevation gradient, pollinator, *Bombus*, mountain, tree line

@@

<!-- This is a paper about mountains. -->

@@

Bumble bees (Hymenoptera: *Bombus* spp.) are classic model organisms in the field of community ecology, but they have more recently become conservation priorities due to well-documented population declines and vulnerability to climate change [@Goulson2008-oe; @Soroye2020-zb]. Bumble bees are characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Upslope movement of bumble bee species in response to climate change has already been documented, and this is expected to cause the declines or extinction of some bumble bee species [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. From a purely geometric perspective, populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer habitat shrinkage [@Dirnbock2011-jf], possibly accompanied by increasing genetic isolation [though see @Fijen2020-zg]. But the outcomes of elevational range shifts on bumble bees also hinge upon the current elevational distributions of bumble bees and their floral resources that will be acted upon by climate change. The upward advance of lowland bumble bee species, for example, may introduce novel competitive pressures on highland species and rewire the pollination networks in which they participate [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm], and elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their floral hosts. When range shifts are considered in this complex biotic context, it is especially important to understand nonlinearities or thresholds in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id], since these could produce dramatic ecological consequences that would not be predicted by geometric considerations alone.

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. In the present study, set along ~1400 m elevation gradient in the German Alps, we investigate elevational patterns in a community of bumble bees and wildflowers at three levels of biological organization. First, we ask how elevation shapes the abundances of bumble bee species and their floral resources. Then, we move from the level of individual species to that of community composition and explore the elevational structure of species turnover (i.e. β-diversity) in bumble bees and flowering plants. Finally, we advance to the level of species interactions, investigating the turnover of bumble- bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques designed to capture nonlinear responses to elevation, with a particular emphasis on the tree line ecotone as a key threshold that structures the biological communities of mountains slopes.

@@

The study was conducted 2010-2012 in Berchtesgaden National Park, located in the Northern Limestone Alps of southeast Germany (47.55°N, 12.92°E). The landscape is composed of mountain pastures mainly surrounded by coniferous forests. We selected 25 study sites (60 x 60 m) on mountain pastures at elevations ranging from 641-2032 meters above sea level (m.s.l.) **(Figure 1)**. Fourteen of these pastures are extensively grazed by cattle or sheep, 3 are mowed for hay production, and 8 have been abandoned throughout the last century and are no longer subject to any human management. Study sites were classified as lying either above or below the tree line based on an examination of Google Earth imagery and corroborated by the authors with field experience at our study sites (K. Kallnik, A. F. Maihoff, A. Claßen). Eight sites were located above the tree line, which generally fell at an elevation of around 1500 m.s.l., consistent with the previous descriptions of the tree line in the Berchtesgadener Alps [@Kostler1970-ya; @Mayer1970-zn].

@@

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date. To represent floral abundance from the perspective of each bumble bee species in our network, we first scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of visited genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection on the estimation of floral resource abundance, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp]. We also opted to treat the cryptic *B. terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished reliably, and we similarly pooled representatives of the parasitic *Psithyrus* subgenus --- *B. barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.

@@

We analyzed the abundance of bumble bees and floral resources using separate hierarchical generalized additive models (HGAMs) fit with the Tweedie distribution family (log link function) [@Wood2011-pg; @Pedersen2019-kf]. We opted in both models to collapse the temporal variation of abundance within sites and focus on elevational patterns. For bumble bees, we modeled *peak* abundance at each site as a smooth response to elevation, with bumble bee species as a smoothing factor and a random intercept term. We also included species:year as a random intercept term because exploratory analysis indicated that species varied differently across years. For floral resources, we modeled *mean* flower cover, weighted per bumble bee species, as a smooth response to elevation, with year and bumble bee species as random intercept terms. Both models are follow the "type I" model form described in @Pedersen2019-kf.

@@

For our analyses of species and interaction β-diversity, we again opted to focus on elevational patterns by pooling samples within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated the species and interaction β-diversity between all pairs of sites and partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (identical species interacting in different ways) [@Novotny2009-my; @Poisot2012-fk]. As an initial analysis, we plotted the relationship between site-wise elevation difference and each metric of β-diversity and verified the significance of the relationship using logistic matrix regression [@Goslee2009-ln], with the difference in number of sampling dates between sites as a covariate to control for potential confounding effects of sampling frequency.

To investigate the relationship between elevation and β-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the β-diversity of bumble bees, flora, and (unpartitioned) interactions. In addition to estimating the relationship between differences in the response variable and differences in the predictor variables, GDM captures the slope of this relationship over the range of each predictor variable, revealing potential variation in the amount of change in the response induced by a given change in a predictor. This enabled us to ask whether the relationship between β-diversity and elevation exhibits thresholds or other nonlinearities. To control for potential confounding effects of geographic proximity and number of sampling dates per site, these terms were added as covariates in the GDM.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. *B. pascuorum* ("pasc"), *B. pratorum* ("prat"), *B. soroeensis* ("soro"), *B. terrestris/lucorum* ("telu"), and *B. wurflenii* ("wurf") --- accounted for the bulk of overall bumble bee abundance, while *B. hortorum* ("hort"), *B. jonellus* ("jone"), *B. psithyrus* ("psit"), *B. monticola* ("mont"), *B. mendax* ("mend"), *B. mucidus* ("muci"), *B. pyrenaeus* ("pyre"), and *B. gerstaeckeri* ("gers") . *B. hypnorum*, *B. humilis*, and *B. lapidarius* were recorded only sporadically and at low abundance, so we omitted them from all analyses.

@@

```{r fig2, echo = FALSE,  fig.cap = paste("Proportional abundances (cumulative within site) of bumble bee species in our study system, ordered by mean proportional abundance.")}
knitr::include_graphics("../output/bb_abundsum.png")
```

@@

The peak abundance of all bumble bee species responded significantly to elevation (p ≤ 0.04) **(Figures 3, 4)**. *B. mendax*, *B. monticola*, *B. mucidus*, and *B. pyrenaeus* exhibited a positive response to elevation and peaked in abundance above the tree line (~ 1500 m.s.l.), consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. The abundance of *B. soroeensis* also tended to increase with elevation and peaked above the tree line, but it was also abundant at lower elevations. In contrast, the abundance of *B. hortorum*, *B. pascorum*, and *B. psithyrus* declined with elevation. This effect was especially pronounced in *B. pascuorum*, which was also the most abundant species in our study system overall. *B. gerstaeckeri*, *B. jonellus*, *B. terrestris-lucorum*, and *B. wurflenii* peaked at mid-elevation below the tree line, while *B. pratorum* was unique in exhibiting a sharp peak of abundance directly *at* the tree line.

```{r fig3, echo = FALSE,  fig.cap = paste("Fitted GAM smooths of bumble bee abundance on elevation, faceted by bumble bee species and plotted over original data. Data points are color-coded to indicate whether they represent sites above (red) or below (blue) the tree line. Note the different scales of the y-axes.")}
knitr::include_graphics("../output/gam_bb_abund.png")
```

```{r fig4, echo = FALSE,  fig.cap = paste("Conditional effect of elevation on the peak abundance of each bumble bee species. Effects are plotted on the link scale, and dashed lines depict 95\\% confidence intervals.")}
knitr::include_graphics("../output/gam_bb_abund_grid.png")
```

Mean floral resource abundance likewise responded significantly (p << 0.0001) to elevation for all bumble bee species **(Figures 5, 6)**. *B. mendax*, *B. monticola*, *B. mucidus*, *B. soroeensis*, and *B. wurflenii* exhibited a weakly to strongly bimodal pattern of flora resource abundance, with peaks around 1000 m.s.l. and 1600 m.s.l., the latter peak corresponding closely to the location of the tree line. *B. jonellus* and *B. pyrenaeus* exhibited a unimodal pattern peaking near 1600 m.s.l. The remaning species --- *B. gerstaeckeri*, *B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. psithyrus*, and *B. terrestris-lucorum* experienced linear or quadratic decline in floral resource abundance with increasing elevation. Floral resource abundance was low for all species at elevations above ~1600 m.s.l.

```{r fig5, echo = FALSE,  fig.cap = paste("Fitted GAM smooths of floral abundance on elevation, faceted by bumble bee species and plotted over original data. Data points are color-coded to indicate whether they represent sites above (red) or below (blue) the tree line. Note the different scales of the y-axes.")}
knitr::include_graphics("../output/gam_fl_abund.png")
```

@@

Floral β-diversity among our study sites was very high overall and responded steeply to elevation difference between sites **(Figure 3A)**. Even sites at similar elevation exhibited ~40% species turnover, and the most widely separated sites (elevation difference > 1250 m) differed by more than 85%.

Bumble bee β-diversity was, in comparison to floral β-diversity, both lower overall and less responsive to elevational difference between sites **(Figure 3A)**. Sites at similar elevation exhibited ~25% species turnover, and species turnover between the most widely separated sites remained less than 50%.

Total interaction β-diversity was >75% between sites at similar elevation and approached perfect dissimilarity in the most widely separated sites **(Figure 3B)**. Partitioning revealed that total interaction β-diversity was driven primarily by species turnover, which accounted for ~50% of total β-diversity for sites at similar elevation and ~90% of total β-diversity for the most widely separated sites. Species turnover consisted mainly of turnover in the floral community **(Figure 3C)**, though the joint turnover of plants and bumble bees accounted for ~25% of total species turnover between the most widely separated sites. Interaction rewiring was most significant between sites at similar elevation, where it accounted for ~30% of total β-diversity, but its share of total β-diversity declined to ~10% in the most widely separated sites **(Figure 3B)**.

Logistic matrix regression confirmed the significance (p < 0.01) of the relationship between each β-diversity metric and elevation **(Table 1)**.

```{r fig7, echo = FALSE,  fig.cap = paste("Species-level (A) and interaction-level (B, C) plotted against elevational difference. Each point represents the β-diversity (or β-diversity partition) between a pair of sites, and lines represent the overall relationship between β-diversity and elevation difference using binomial regression smooths. Standard errors are not plotted because they would be misleading due to the non-independence inherent to distance matrix regression, but all regressions were significant (p < 0.01). Interaction β-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned β-diversity, ST = β-diversity due to species turnover, OS =  β-diversity due to interaction rewiring, ST.h =  β-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = β-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  β-diversity due to joint species turnover in higher and lower trophic levels.")}
knitr::include_graphics("../output/beta_stack.png")
```

>>>>

---
title: "The tree line ecotone punctuates the elevational turnover of plants, pollinators, and their interactions"

author: Douglas B. Sponsler^1^*, Fabrice Requier^2^, Katharina Kallnik^1^, Alice Claßen^1^, A. Fabienne Maihoff^1^, Johanna Sieger^3^, Ingolf Steffan-Dewenter^1^

@@

^3^ Naturpark Frankenhöhe, Ansbach, Germany

@@

# Open Research Statement

Data are not yet provided but will be permanently archived on Dryad if the paper is accepted for publication. This submission uses novel code, which is provided as Supporting Information to be evaluated as part of the submission. If the paper is accepted, the code will be archived on Dryad alongside our raw data.

@@

The structuring of biological communities along mountain slopes is complex, and elevational range shifts in response to climate change involve more than merely tracking suitable temperature envelopes. Bumble bees (Hymentopera: *Bombus* spp.) are cold-adapted species associated with mountain habitats, and they are already exhibiting upslope range shifts that are expected to result in habitat loss, novel competitive interactions, and the rewiring of pollination networks. Interpreting these shifts, however, requires an understanding of the current elevational patterns of bumble bees and the floral mutualists that are being acted upon by climate change. We recorded bumble- bee-flower interactions over three years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling, we analyze the elevational patterns at the nested levels of species abundance, species β-diversity, and interaction β-diversity. We demonstrate that the tree line ecotone is (1) a distributional interface between low/mid- and high-elevation bumble bee species, (2) a threshold above which floral resource availability sharply decreases, and (3) a zone of accelerated turnover of floral composition and bumble- bee-flower interactions. Our results show that linear elevational temperature gradients are ecologically punctuated, and the outcomes of climate-induced range shifts may hinge upon dynamics at the tree line ecotone.

<!-- The structuring of biological communities along mountain slopes is complex, and the elevational range shifts that are now occurring in response to climate change involve more than merely tracking suitable temperature envelopes. When species move, they do so in the context of biological communities, and the outcomes of these movements depend on how and to what extent biotic interactions are reordered. Bumble bees (Hymentopera: *Bombus* spp.) are cold-adapted species associated with mountain habitats, and they have already exhibited measurable upslope range shifts that are expected to result in habitat loss, novel competitive interactions, and the rewiring of pollination networks. Predicting and interpreting these shifts, however, requires an understanding of the current elevational patterns of bumble bees and the floral mutualists that are being acted upon by climate change. In this study, we sampled bumble bees, flowering plants, and their interactions over three consecutive years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling techniques, we analyze the elevational patterns of this community at three levels of biological organization: species abundance, species β-diversity, and interaction β-diversity. We demonstrate complex, nonlinear responses to elevation at all three levels of organization. In particular, we identify the tree line ecotone as (1) a distributional interface between low/mid- and high-elevation bumble bee species, (2) a threshold above which floral resource availability sharply decreases, and (3) a zone of accelerated turnover of flowering plants and bumble- bee-flower interactions. The implications of these findings extend beyond the particular case of bumble bees to demonstrate that linear elevational temperature gradients are ecologically punctuated, and the outcomes of climate-induced range shifts with depend on how species and species-interactions change as they cross --- or fail to cross --- thresholds like the tree line ecotone. -->

*Keywords*: climate, beta-diversity, elevation gradient, *Bombus*, bumble bee, mountain ecology, pollination network

@@

@@

Bumble bees (Hymenoptera: *Bombus* spp.) are classic model organisms in the field of community ecology, but they have more recently become conservation priorities due to well-documented population declines and vulnerability to climate change [@Goulson2008-oe; @Soroye2020-zb]. Bumble bees are characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Upslope movement of bumble bee species in response to climate change has already been documented, and this is expected to cause the declines or extinction of some bumble bee species [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; Biella2017-ni; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. From a purely geometric perspective, populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer habitat shrinkage [@Dirnbock2011-jf], possibly accompanied by increasing genetic isolation [though see @Fijen2020-zg]. But the outcomes of elevational range shifts on bumble bees also hinge upon the current elevational distributions of bumble bees and their floral resources that will be acted upon by climate change. The upward advance of lowland bumble bee species, for example, may introduce novel competitive pressures on highland species and rewire the pollination networks in which they participate [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm], and elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their floral hosts. When range shifts are considered in this complex biotic context, it is especially important to understand nonlinearities or thresholds in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id], since these could produce dramatic ecological consequences that would not be predicted by geometric considerations alone.

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic process of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. In the present study, set along ~1400 m elevation gradient in the German Alps, we investigate elevational patterns in a community of bumble bees and wildflowers at three levels of biological organization. First, we ask how elevation shapes the abundances of bumble bee species and their floral resources. Then, we move from the level of individual species to that of community composition and explore the elevational structure of species turnover (i.e. β-diversity) in bumble bees and flowering plants. Finally, we advance to the level of species interactions, investigating the turnover of bumble- bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques designed to capture nonlinear responses to elevation, with a particular emphasis on the tree line ecotone as a key threshold that structures the biological communities of mountains slopes.

@@

The study was conducted 2010-2012 in Berchtesgaden National Park, located in the Northern Limestone Alps of southeast Germany (47.55°N, 12.92°E). The landscape is composed of mountain pastures mainly surrounded by coniferous forests. We selected 25 study sites (60 x 60 m) on mountain pastures at elevations ranging from 641-2032 meters above sea level (m.s.l.) **(Figure 1)**. Fourteen of these pastures are extensively grazed by cattle or sheep, 3 are mowed for hay production, and 8 have been abandoned throughout the last century and are no longer subject to any human management. Study sites were classified as lying either above or below the tree line based on an examination of Google Earth imagery corroborated by field experience at our study sites. Eight sites were located above the tree line, which generally fell at an elevation of around 1500 m.s.l., consistent with the previous descriptions of the tree line in the Berchtesgadener Alps [@Kostler1970-ya; @Mayer1970-zn].

@@

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date. To weight floral abundance based on the foraging preferences of each bumble bee species in our network, we scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of *visited* genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection on the estimation of floral resource abundance, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp]. We also opted to treat the cryptic *B. terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished, and we similarly pooled representatives of the parasitic *Psithyrus* subgenus --- *B. barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.

@@

We analyzed the abundance of bumble bees and floral resources using separate hierarchical generalized additive models (HGAMs) [@Wood2011-pg; @Pedersen2019-kf]. In both models, we opted to collapse the temporal variation of abundance within sites and focus on elevational patterns. For bumble bees, we modeled *peak* abundance at each site as a smooth response to elevation, with bumble bee species as a smoothing factor, management (grazing, mowing, or unmanaged) as a fixed effect, and species:year as a random intercept term (exploratory analysis indicated that species varied differently across years). For floral resources, we modeled *mean* flower cover, weighted per bumble bee species (i.e. the sum of *visited* genera), as a smooth response to elevation, with management as a fixed effect, and year and bumble bee species as random intercept terms. For the bumble bee model, we used the negative binomial distribution family (log link function), and for the floral model we used a Tweedie family distribution (log link function). Both models follow the "type I" model form described in @Pedersen2019-kf.

@@

For our analyses of species and interaction β-diversity, we again opted to focus on elevational patterns by pooling samples within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated the species and interaction β-diversity between all pairs of sites and partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (interaction turnover despite species constancy) [@Novotny2009-my; @Poisot2012-fk]. As an initial analysis, we plotted the relationship between site-wise elevation difference and each metric of β-diversity and verified the significance of the relationship using logistic matrix regression [@Goslee2009-ln], with the difference in number of sampling dates between sites as a covariate to control for potential confounding effects of sampling frequency.

To investigate the relationship between elevation and β-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the β-diversity of bumble bees, flora, and (unpartitioned) interactions. In addition to estimating the relationship between differences in the response variable and differences in the predictor variables, GDM captures the slope of this relationship over the range of each predictor variable, revealing potential variation in the amount of change in the response induced by a given change in a predictor. This enabled us to ask whether the relationship between β-diversity and elevation exhibits thresholds or other nonlinearities. To control for potential confounding effects of geographic proximity and number of sampling dates per site, these terms were added as covariates in the GDM. GDM is designed to model continuous predictors, and while the inclusion of categorical variables is theoretically possible [@Ferrier2007-ji],

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. *B. pascuorum* ("pasc"), *B. pratorum* ("prat"), *B. soroeensis* ("soro"), *B. terrestris/lucorum* ("telu"), and *B. wurflenii* ("wurf") --- accounted for the bulk of overall bumble bee abundance, while *B. hortorum* ("hort"), *B. jonellus* ("jone"), *B. psithyrus* ("psit"), *B. monticola* ("mont"), *B. mendax* ("mend"), *B. mucidus* ("muci"), *B. pyrenaeus* ("pyre"), and *B. gerstaeckeri* ("gers") . *B. hypnorum*, *B. humilis*, and *B. lapidarius* were recorded only sporadically and at low abundance, so we omitted them from all analyses. See Appendix X for a summary of species-wise abundances.

@@

@@

The peak abundance of all bumble bee species responded significantly to elevation (p ≤ 0.04) **(Figure 2)**. *B. mendax*, *B. monticola*, *B. mucidus*, and *B. pyrenaeus* exhibited a positive response to elevation and peaked in abundance above the tree line (~ 1500 m.s.l.), consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. The abundance of *B. soroeensis* also tended to increase with elevation and peaked above the tree line, but it was also abundant at lower elevations. In contrast, the abundance of *B. hortorum*, *B. pascorum*, and *B. psithyrus* declined with elevation. This effect was especially pronounced in *B. pascuorum*, likely because its exceptionally long nesting cycle is incompatible with the short growing season at high elevations [@Neumayer2009-gx]. *B. gerstaeckeri*, *B. jonellus*, *B. terrestris-lucorum*, and *B. wurflenii* peaked at mid-elevation below the tree line, while *B. pratorum* was unique in exhibiting a sharp peak of abundance directly *at* the tree line.

<!-- ```{r fig3, echo = FALSE,  fig.cap = paste("Fitted GAM smooths of bumble bee abundance on elevation, faceted by bumble bee species and plotted over original data. Data points are color-coded to indicate whether they represent sites above (red) or below (blue) the tree line. Note the different scales of the y-axes.")} -->
<!-- knitr::include_graphics("../output/gam_bb_abund.png") -->
<!-- ``` -->

```{r fig2, echo = FALSE,  fig.cap = paste("Conditional effect of elevation on the peak abundance of each bumble bee species. Effects are plotted on the link scale, and dashed lines depict 95\\% confidence intervals.")}
knitr::include_graphics("../output/gam_bb_abund_grid.png")
```

Mean floral resource abundance likewise responded significantly (p << 0.0001) to elevation for all bumble bee species **(Figure 3)**. *B. mendax*, *B. monticola*, *B. mucidus*, *B. soroeensis*, and *B. wurflenii* exhibited a weakly to strongly bimodal pattern of flora resource abundance, with peaks around 1000 m.s.l. and 1600 m.s.l., the latter peak corresponding closely to the location of the tree line. *B. jonellus* and *B. pyrenaeus* exhibited a unimodal pattern peaking near 1600 m.s.l. The remaining species --- *B. gerstaeckeri*, *B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. psithyrus*, and *B. terrestris-lucorum* experienced linear or quadratic decline in floral resource abundance with increasing elevation. Floral resource abundance was low for all species at elevations above ~1600 m.s.l.

# ```{r fig5, echo = FALSE,  fig.cap = paste("Fitted GAM smooths of floral abundance on elevation, faceted by bumble bee species and plotted over original data. Data points are color-coded to indicate whether they represent sites above (red) or below (blue) the tree line. Note the different scales of the y-axes.")}
# knitr::include_graphics("../output/gam_fl_abund.png")
# ```

@@

Floral β-diversity among our study sites was very high overall and responded steeply to elevation difference between sites **(Figure 4A)**. Even sites at similar elevation exhibited ~40% species turnover, and the most widely separated sites (elevation difference > 1250 m) differed by more than 85%.

Bumble bee β-diversity was, in comparison to floral β-diversity, both lower overall and less responsive to elevational difference between sites **(Figure 4A)**. Sites at similar elevation exhibited ~25% species turnover, and species turnover between the most widely separated sites remained less than 50%.

Total interaction β-diversity was >75% between sites at similar elevation and approached perfect dissimilarity in the most widely separated sites **(Figure 4B)**. Partitioning revealed that total interaction β-diversity was driven primarily by species turnover, which accounted for ~50% of total β-diversity for sites at similar elevation and ~90% of total β-diversity for the most widely separated sites. Species turnover consisted mainly of turnover in the floral community **(Figure 4C)**, though the joint turnover of plants and bumble bees accounted for ~25% of total species turnover between the most widely separated sites. Interaction rewiring was most significant between sites at similar elevation, where it accounted for ~30% of total β-diversity, but its share of total β-diversity declined to ~10% in the most widely separated sites **(Figure 4B)**.

Logistic matrix regression confirmed the significance (p < 0.01) of the relationship between each β-diversity metric and elevation.

```{r fig4, echo = FALSE,  fig.cap = paste("Species-level (A) and interaction-level (B, C) plotted against elevational difference. Each point represents the β-diversity (or β-diversity partition) between a pair of sites, and lines represent the overall relationship between β-diversity and elevation difference using binomial regression smooths. Standard errors are not plotted because they would be misleading due to the non-independence inherent to distance matrix regression, but all regressions were significant (p < 0.01). Interaction β-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned β-diversity, ST = β-diversity due to species turnover, OS =  β-diversity due to interaction rewiring, ST.h =  β-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = β-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  β-diversity due to joint species turnover in higher and lower trophic levels.")}
knitr::include_graphics("../output/beta_stack.png")
```
